---
title: "contingency_table"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{contingency_table}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(contingencytable)
```


This is a package dedicated to doing simple, two-way contingency table analysis. Too many packages out there to do the simple things we all need to do, even it is just preliminary analysis...

Let's start with... loading the package! 

```{r}
library(contingencytable)
```
... and generating a random contingency table!

```{r}
X = matrix(sample(1:50,12),4,3)

print(X)
```

A two-way contingency table is a contingency table relating two variables, X and Y. It is represented as an I $\times$ J table, with count $n_{ij}$ being the number of subjects with $X = i$ and $Y = j$. The total number of subjects observed ois $n = \sum_i \sum_j n_{ij}$. This is the observed contingency table. What we are interested in is the population level contingency table, that is also I $\times$ J and $\pi_{ij}$ is the true proportion of the population with $X = i$ and $Y = j$. Here,   $\sum_i \sum_j \pi_{ij} = 1$. 

We use the cell frequencies $$\hat{\pi}_{ij} = n_{ij}/n$$ to estimate $\pi_{ij}$. We want to test independence... so we want to check $H_0 : \pi_{ij} = \pi_{+j}\pi_{i+}$, where $\pi_{+j}$ and $\pi_{i+}$ are both marginal over the rows and the columns, respectively. So $\pi_{+j} = P(Y = j) = \sum_i \pi_{ij}$ and $\pi_{i+} = P(X = i) = \sum_j \pi_{ij}$. To test this alternative hypothesis, we present a few options in contingency table. 

## Pearson Chi- Squared

Under the null, we should expect cell frequencies to be $n \hat{\pi}_{ij} = n \hat{\pi}_{i+}\hat{\pi}_{+j} = n_{+j}n_{i+}/n = \hat{\mu_{ij}}$. We don't observe those frequencies, however, but rather we observe $n_{ij}$. The Pearson Statistic is given by $$X^2 = \displaystyle\sum_{i =1}^I \sum_{j = 1}^J \frac{(n_{ij}-\hat{\mu_{ij}})^2}{\hat{\mu_{ij}}}$$. This statistc is distrbuted $\chi ^2_{(I-1)\times(J-1)}$ asymptotically under the null hypothesis. 

```{r setup}
analysis = contingency_table(X)

analysis$chi2

analysis$p_chi2

```

So we can see that there is mild significance to say X and Y, in this contingency table, are associated! 

## Likelihood Ratio Statistic 

```{r}
analysis$g2

analysis$p_g2
```
